\section*{Exercises}


\begin{ex} In the following, polar coordinates $(r,\theta) $ for a
point in the plane are given. Find the corresponding Cartesian coordinates.

\begin{enumerate}
\item $(2,\pi /4) $

\item $(-2,\pi /4) $

\item $(3,\pi /3) $

\item $(-3,\pi /3) $

\item $(2,5\pi /6) $

\item $(-2,11\pi /6) $

\item $(2,\pi /2) $

\item $(1,3\pi /2) $

\item $(-3,3\pi /4) $

\item $(3,5\pi /4) $

\item $(-2,\pi /6) $
\end{enumerate}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the following Cartesian coordinates $(x,y)$. Find polar coordinates corresponding to these points.

\begin{enumerate}
\item $(-1,1) $

\item $(\sqrt{3},-1) $

\item $(0,2) $

\item $(-5,0) $

\item $(-2\sqrt{3},2) $

\item $(2,-2) $

\item $(-1,\sqrt{3}) $

\item $(-1,-\sqrt{3}) $
\end{enumerate}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} The following relations are written in terms of Cartesian coordinates $(x, y)$. Rewrite them in terms of polar coordinates, $(r, \theta)$.

\begin{enumerate}
\item $y=x^{2}$

\item $y=2x+6$

\item $x^{2}+y^{2}=4$

\item $x^{2}-y^{2}=1$
\end{enumerate}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Use a calculator or computer algebra system to graph the following
polar relations.

\begin{enumerate}
\item $r=1-\sin (2\theta) ,\theta \in \mat{0,2\pi} $

\item $r=\sin (4\theta) ,\theta \in \mat{0,2\pi} $

\item $r=\cos (3\theta) +\sin (2\theta) ,\theta
\in \mat{0,2\pi} $

\item $r=\theta ,\ \theta \in \mat{0,15} $
\end{enumerate}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Graph the polar equation $r=1+\sin \theta $ for $\theta \in \mat{0,2\pi}$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Graph the polar equation $r=2+\sin \theta $ for $\theta \in \mat{0,2\pi}$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Graph the polar equation $r=1+2\sin \theta $ for $\theta \in \mat{0,2\pi}$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Graph the polar equation $r=2+\sin (2\theta) $ for $\theta \in \mat{0,2\pi
}$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Graph the polar equation $r=1+\sin (2\theta) $ for $\theta \in \mat{0,2\pi
}$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Graph the polar equation $r=1+\sin (3\theta) $ for $\theta \in \mat{0,2\pi
}$.
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} Describe how to solve for $r$ and $\theta $ in terms of $x$ and $y$ in polar
coordinates.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} This problem deals with parabolas, ellipses, and
hyperbolas and their equations. Let $l,e>0$ and consider
\begin{equation*}
r=\frac{l}{1\pm e\cos \theta }
\end{equation*}
Show that if $e=0$, the graph of this equation gives a circle. Show that if $0<e<1$, the graph is an ellipse, if $e=1$
it is a parabola and if $e>1$, it is a hyperbola.
%\begin{sol}
%\end{sol}
\end{ex}

