\section*{Exercises}

\begin{ex}
  Which of the following linear transformations $T:\C^2\to\C^2$ is unitary?
  \begin{enumerate}
  \item $T\paren{\begin{mymatrix}{c} x \\ y \end{mymatrix}}
    = \begin{mymatrix}{c} x \\ iy \end{mymatrix}$.
  \item $T\paren{\begin{mymatrix}{c} x \\ y \end{mymatrix}}
    = \begin{mymatrix}{c} iy \\ -ix \end{mymatrix}$.
  \item $\def\arraystretch{1.4}
    T\paren{\begin{mymatrix}{c} x \\ y \end{mymatrix}}
    = \begin{mymatrix}{c}
      x + iy \\
      ix + y \\
    \end{mymatrix}$.
  \item $\def\arraystretch{1.3}
    T\paren{\begin{mymatrix}{c} x \\ y \end{mymatrix}}
    = \begin{mymatrix}{c}
      \frac{1}{\sqrt{2}} x + \frac{i}{\sqrt{2}} y \\
      \frac{1}{\sqrt{2}} x - \frac{i}{\sqrt{2}} y \\
    \end{mymatrix}$.
  \end{enumerate}
  \begin{sol}
    (a) Yes. (b) Yes. (c) No. (d) Yes.
  \end{sol}
\end{ex}

\begin{ex}
  Determine which of the following matrices are unitary.
  \begin{equation*}
    A = \begin{mymatrix}{cc}
      0 & i \\
      i & 0 \\
    \end{mymatrix},
    \quad
    B = \frac{1}{5}
    \begin{mymatrix}{cc}
      4 & 3i \\
      3 & 4i \\
    \end{mymatrix},
    \quad
    C = \frac{1}{\sqrt{5}}
    \begin{mymatrix}{cc}
      1 & 2i \\
      2i & 1 \\
    \end{mymatrix}.
  \end{equation*}
  \begin{sol}
    (a) Yes. (b) No. (c) Yes.
  \end{sol}
\end{ex}

\begin{ex}
  Determine which of the following matrices are unitary.
  \begin{equation*}
    A = \frac{1}{2}
    \begin{mymatrix}{ccc}
      1 & i & 0 \\
      0 & 1 & i \\
      i & 0 & 1 \\
    \end{mymatrix},
    \quad
    B = \frac{1}{3}
    \begin{mymatrix}{ccc}
      1   & 2  & 2i \\
      -2i & 2i & 1  \\
      2i  & i  & 2  \\
    \end{mymatrix},
    \quad
    C = \frac{1}{2}
    \begin{mymatrix}{cccc}
      1 &  1 &  1 &  1 \\
      1 &  i & -1 & -i \\
      1 & -1 &  1 & -1 \\
      1 & -i & -1 &  i \\
    \end{mymatrix}.
  \end{equation*}
  \begin{sol}
    (a) No. (b) Yes. (c) Yes.
  \end{sol}
\end{ex}

\begin{ex}
  Assume $A$ and $B$ are unitary $3\times 3$-matrices. Then which of
  the following matrices are unitary?
  \begin{equation*}
    (a)~ AB\quad
    (b)~ A+B\quad
    (c)~ iA\quad
    (d)~ -A\quad
    (e)~ 2A\quad
    (f)~ B^{-1}\quad
    (g)~ A^T\quad
    (h)~ A^*\quad
    (i)~ A^2\quad
    (j)~ ABA^{-1}
  \end{equation*}
  \begin{sol}
    All except (b) and (e) are unitary.
  \end{sol}
\end{ex}

\begin{ex}
  Assume $A$ and $B$ are $3\times 3$-matrices, and assume $A$ is
  unitary and $B$ is invertible. Then which of the following matrices
  are unitary?
  \begin{equation*}
    (a)~ AB\quad
    (b)~ A+B\quad
    (f)~ BAB^{-1}\quad
    (g)~ ABA^{-1}\quad
    (h)~ BB^*\quad
    (i)~ BB^{-1}
  \end{equation*}
  \begin{sol}
    Only (f) is unitary.
  \end{sol}
\end{ex}

\begin{ex}
  Which of the following matrices are hermitian? Which ones are
  symmetric?
  \begin{equation*}
    A = \begin{mymatrix}{cc} 0 & 1 \\ 1 & 0 \end{mymatrix}, \quad
    B = \begin{mymatrix}{cc} i & 2 \\ 2 & -i \end{mymatrix}, \quad
    C = \begin{mymatrix}{cc} 0 & i \\ -i & 0 \end{mymatrix}, \quad
    D = \begin{mymatrix}{cc} 1 & 1+i \\ 1-i & 2 \end{mymatrix}.
  \end{equation*}
\end{ex}

\begin{ex}
  Which of the following matrices are hermitian? Which ones are
  symmetric?
  \begin{equation*}
    A = \begin{mymatrix}{ccc}
      0   & -2 & 1+i \\
      -2  &  3 & i   \\
      1-i &  i & 2   \\
    \end{mymatrix}, \quad
    B = \begin{mymatrix}{ccc}
      3     & i    & -1+2i \\
      -i    & 0    & -1+i  \\
      -1-2i & -1-i & 2     \\
    \end{mymatrix}, \quad
    C = \begin{mymatrix}{ccc}
      i & i & i \\
      i & i & i \\
      i & i & i \\
    \end{mymatrix}.
  \end{equation*}
\end{ex}

\begin{ex}
  Assume $A$ and $B$ are hermitian $3\times 3$-matrices, and $A$ is
  invertible. Then which of the following matrices are hermitian?
  \begin{equation*}
    (a)~ AB\quad
    (b)~ A+B\quad
    (c)~ iA\quad
    (d)~ -A\quad
    (e)~ 2A\quad
    (f)~ A^*BA\quad
    (g)~ A^{-1}BA\quad
    (h)~ BB^*\quad
    (i)~ A^2
  \end{equation*}
  \begin{sol}
    All except (a), (c), and (d) are hermitian.
  \end{sol}
\end{ex}

\begin{ex}
  Unitarily diagonalize the hermitian matrix
  $A=\begin{mymatrix}{cc}
    2  & 1+i \\
    1-i & 3 \\
  \end{mymatrix}$.
  \begin{sol}
    Eigenvalues: $\eigenvar_1=1$ and $\eigenvar_2=4$. Normalized
    eigenvectors:
    $\vect{v}_1 = \frac{1}{\sqrt{3}}\begin{mymatrix}{c} 1+i \\ -1 \end{mymatrix}$,
    $\vect{v}_2 = \frac{1}{\sqrt{3}}\begin{mymatrix}{c} 1 \\ 1-i \end{mymatrix}$.
    \begin{equation*}
      D = \begin{mymatrix}{cc} 1 & 0 \\ 0 & 4 \end{mymatrix}, \quad
      P = \frac{1}{\sqrt{3}}\begin{mymatrix}{cc} 1+i & 1 \\ -1 & 1-i \end{mymatrix}.
    \end{equation*}
  \end{sol}
\end{ex}

\begin{ex}
  Unitarily diagonalize the hermitian matrix
  $A=\begin{mymatrix}{cc}
    0  & -1+2i \\
    -1-2i & 4 \\
  \end{mymatrix}$.
\end{ex}

\begin{ex}
  Unitarily diagonalize the hermitian matrix
  $A=\begin{mymatrix}{ccc}
    2   & 1+i & 1-i  \\
    1-i & 3   & -2i  \\
    1+i & 2i  &   3  \\
  \end{mymatrix}$.
\end{ex}
