\section*{Exercises}

\begin{ex}
  Consider the following collection of points in $\R^2$:
  \begin{equation*}
    \set{
      \begin{mymatrix}{r} 1 \\ -1 \end{mymatrix},
      \begin{mymatrix}{r} -2 \\ 1 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ -2 \end{mymatrix},
      \begin{mymatrix}{r} 2 \\ -2 \end{mymatrix},
      \begin{mymatrix}{r} -2 \\ 1 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ -1 \end{mymatrix},
      \begin{mymatrix}{r} -3 \\ 0 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ 3 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ 0 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ -2 \end{mymatrix}
    }.
  \end{equation*}
  Find the 1-dimensional subspace that best approximates this
  collection of points. What is the total squared distance of the
  points to the subspace? Sketch the subspace and the points.
\end{ex}

\begin{ex}
  Consider the following collection of points in $\R^3$:
  \begin{equation*}
    \set{
      \begin{mymatrix}{r} 0 \\ 0 \\ 9 \end{mymatrix},
      \begin{mymatrix}{r} 1 \\ 0 \\ -3 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ 0 \\ 1 \end{mymatrix}
      \begin{mymatrix}{r} 0 \\ 5 \\ -4 \end{mymatrix},
      \begin{mymatrix}{r} 6 \\ -2 \\ 13 \end{mymatrix},
      \begin{mymatrix}{r} 1 \\ 0 \\ 0 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ 2 \\ 0 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ 1 \\ 1 \end{mymatrix},
      \begin{mymatrix}{r} 2 \\ 0 \\ 0 \end{mymatrix},
      \begin{mymatrix}{r} 4 \\ 0 \\ 0 \end{mymatrix},
      \begin{mymatrix}{r} 1 \\ -3 \\ 0 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ 0 \\ 4 \end{mymatrix}
    }.
  \end{equation*}
  \begin{enumerate}
  \item Find the 1-dimensional subspace that best approximates this
    collection of points.
  \item Find the 2-dimensional subspace that best approximates this
    collection of points.
  \item What is the 3-dimensional subspace that best approximates this
    collection of points?
  \end{enumerate}
  In each case, what is the total squared distance of the points to
  the subspace?
\end{ex}

\begin{ex}
  Find the principal components of the matrix
  \begin{equation*}
    A = \begin{mymatrix}{ccccccccc}
      0 & 0 & 0 & 0 & -1 & 0 & 3 & 0 & 1 \\
      0 & 1 & 0 & 0 & 2 & 3 & 0 & 2 & -1 \\
      1 & -1 & -3 & 3 & -1 & 0 & 0 & -2 & 2 \\
    \end{mymatrix}.
  \end{equation*}
\end{ex}

\begin{ex}
  Compute the centroid of the following collection of points:
  \begin{equation*}
    \set{
      \begin{mymatrix}{r} 2 \\ -1 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ 1 \end{mymatrix},
      \begin{mymatrix}{r} 4 \\ -3 \end{mymatrix},
      \begin{mymatrix}{r} -4 \\ -4 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ -2 \end{mymatrix},
      \begin{mymatrix}{r} -2 \\ -1 \end{mymatrix},
      \begin{mymatrix}{r} 0 \\ -2 \end{mymatrix},
      \begin{mymatrix}{r} -2 \\ -4 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ -1 \end{mymatrix},
      \begin{mymatrix}{r} -5 \\ -5 \end{mymatrix}
    }.
  \end{equation*}
\end{ex}

\begin{ex}
  Consider the following collection of points in $\R^2$:
  \begin{equation*}
    \set{
      \begin{mymatrix}{r} 4 \\ -3 \end{mymatrix},
      \begin{mymatrix}{r} 10 \\ -9 \end{mymatrix},
      \begin{mymatrix}{r} 4 \\ -7 \end{mymatrix},
      \begin{mymatrix}{r} -2 \\ 3 \end{mymatrix},
      \begin{mymatrix}{r} 10 \\ -3 \end{mymatrix},
      \begin{mymatrix}{r} 4 \\ -3 \end{mymatrix},
      \begin{mymatrix}{r} 5 \\ -3 \end{mymatrix},
      \begin{mymatrix}{r} 4 \\ 1 \end{mymatrix},
      \begin{mymatrix}{r} -2 \\ -3 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ -3 \end{mymatrix}
    }.
  \end{equation*}
  Compute the centroid, and then find the 1-dimensional affine
  subspace that best approximates this collection of points. What is
  the total squared distance of the points to the subspace?
\end{ex}

\begin{ex}
  Consider the following collection of points in $\R^3$:
  \begin{equation*}\scriptstyle
    \set{
      \begin{mymatrix}{r} 0 \\ -2 \\ 2 \end{mymatrix},
      \begin{mymatrix}{r} 6 \\ 2 \\ 2 \end{mymatrix},
      \begin{mymatrix}{r} 2 \\ 2 \\ 6 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ 2 \\ 2 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ 0 \\ 5 \end{mymatrix},
      \begin{mymatrix}{r} 5 \\ 0 \\ 2 \end{mymatrix},
      \begin{mymatrix}{r} 4 \\ -2 \\ -2 \end{mymatrix},
      \begin{mymatrix}{r} 5 \\ 0 \\ 2 \end{mymatrix},
      \begin{mymatrix}{r} 1 \\ 0 \\ 2 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ 0 \\ -1 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ 0 \\ 1 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ 0 \\ 3 \end{mymatrix},
      \begin{mymatrix}{r} 1 \\ 0 \\ 2 \end{mymatrix},
      \begin{mymatrix}{r} 3 \\ -2 \\ 2 \end{mymatrix}
    }.
  \end{equation*}
  Find the 1- and 2-dimensional affine subspaces that best approximate
  this collection of points. What is the total squared
  distance of the points to each subspace?
\end{ex}
