\section*{Exercises}


\begin{ex}
  Which of the following sets are subspaces of $\R^3$? Explain.
  \begin{enumerate}
  \item $V_1=\set{\left.\vect{u}=\begin{mymatrix}{c}
          u_{1} \\ u_{2} \\ u_{3}
        \end{mymatrix}~\right\vert~ \abs{u_{1}} \leq 4}$.
  \item $V_2=\set{\left.\vect{u}=\begin{mymatrix}{c}
          u_{1} \\ u_{2} \\ u_{3}
        \end{mymatrix}~\right\vert~ u_{i}\geq 0\text{ for each }i=1,2,3}$.
  \item $V_3=\set{\left.\vect{u}=\begin{mymatrix}{c}
          u_{1} \\ u_{2} \\ u_{3}
        \end{mymatrix}~\right\vert~ u_{3}+u_{1}=2u_{2}}$.
  \item $V_4=\set{\left.\vect{u}=\begin{mymatrix}{c}
          u_{1} \\ u_{2} \\ u_{3}
        \end{mymatrix}~\right\vert~ u_{3}\geq u_{1}}$.
  \item $V_5=\set{\left.\vect{u}=\begin{mymatrix}{c}
          u_{1} \\ u_{2} \\ u_{3}
        \end{mymatrix}~\right\vert~ u_{3}=u_{1}=0}$.
  \end{enumerate}

  \begin{sol}
    \begin{enumerate}
    \item
      No. We have $\begin{mymatrix}{r}
        1 \\ 0 \\ 0 \\ 0
      \end{mymatrix} \in V_1$ but $10\begin{mymatrix}{r}
        1 \\ 0 \\ 0 \\ 0
      \end{mymatrix} \notin V_1$.
    \item
      This is not a subspace. The vector $\begin{mymatrix}{r}
        1 \\ 1 \\ 1 \\ 1
      \end{mymatrix} $
      is in $V_2$. However, $(-1) \begin{mymatrix}{r}
        1 \\ 1 \\ 1 \\ 1
      \end{mymatrix} $ is not.
    \item This is a subspace. It contains the zero vector and is
      closed with respect to vector addition and scalar
      multiplication.
    \item
      This
      is not a subspace. The vector $\begin{mymatrix}{r}
        0 \\ 0 \\ 1 \\ 0
      \end{mymatrix} $ is in $V_4$. However $(-1) \begin{mymatrix}{r}
        0 \\ 0 \\ 1 \\ 0
      \end{mymatrix}  = \begin{mymatrix}{r}
        0 \\ 0 \\ -1 \\ 0
      \end{mymatrix} $ is not.
    \item This is a subspace. It contains the zero vector and is
      closed with respect to vector addition and scalar
      multiplication.
    \end{enumerate}
  \end{sol}
\end{ex}

\begin{ex}
  Let $\vect{w}\in \R^{4}$ be a given fixed vector. Let
  \begin{equation*}
    M=\set{\left.\vect{u}
        =\begin{mymatrix}{c}
          u_{1} \\ u_{2} \\ u_{3} \\ u_{4}
        \end{mymatrix} \in \R^{4}~\right\vert~ \vect{w}\dotprod \vect{u}
      =0}.
  \end{equation*}
  Is $M$ a subspace of $\R^4$? Explain.
  \begin{sol}
    Yes, this is a subspace because it contains the zero vector and is
    closed with respect to vector addition and scalar
    multiplication. For example, if $\vect{u},\vect{v}\in M$, then
    $\vect{w}\dotprod\vect{u}=0$ and $\vect{w}\dotprod\vect{v}=0$,
    therefore $\vect{w}\dotprod(\vect{u}+\vect{v})=0$, therefore
    $\vect{u}+\vect{v}\in M$.
  \end{sol}
\end{ex}

\begin{ex}
  Let $\vect{w},\vect{v}$ be given vectors in $\R^{4}$ and define
  \begin{equation*}
    M=\set{\left.\vect{u}=\begin{mymatrix}{c}
        u_{1} \\ u_{2} \\ u_{3} \\ u_{4}
      \end{mymatrix} \in \R
      ^{4}~\right\vert~ \vect{w}\dotprod \vect{u}=0\text{ and }\vect{v}\dotprod \vect{u}=0}
    .
  \end{equation*}
  Is $M$ a subspace of $\R^4$? Explain.
  \begin{sol}
    Yes, this is a subspace.
  \end{sol}
\end{ex}

\begin{ex}
  Suppose $V, W$ are subspaces of $\R^{n}$. Let $V\cap W$ be the set
  of all vectors that are in both $V$ and $W$. Show that $V\cap W$
  is also a subspace.
  \begin{sol}
    Because $\vect{0}\in V$ and $\vect{0}\in W$, we have
    $\vect{0}\in V\cap W$. To show that $V\cap W$ is closed under
    addition, assume $\vect{u},\vect{v}\in V\cap W$. Then
    $\vect{u},\vect{v}\in V$, and since $V$ is a subspace we have
    $\vect{u}+\vect{v}\in V$. Similarly $\vect{u},\vect{v}\in W$, and
    since $W$ is a subspace we have $\vect{u}+\vect{v}\in W$. It
    follows that $\vect{u}+\vect{v}$ is in both $V$ and $W$, and
    therefore in $V\cap W$. To show that $V\cap W$ is closed under
    scalar multiplication, assume $\vect{u}\in V\cap W$ and
    $k\in\R$. Then $\vect{u}\in V$, and therefore $k\vect{u}\in
    V$. Similarly $k\vect{u}\in W$, and therefore also
    $k\vect{u}\in V\cap W$.
  \end{sol}
\end{ex}

\begin{ex}
  Let $V$ be a subset of $\R^n$. Show that $V$ is a subspace if and
  only if it is non-empty and the following condition holds: for all
  $\vect{u},\vect{v}\in V$ and all scalars $a,b\in\R$,
  \begin{equation*}
    a\vect{u} + b\vect{v} \in V.
  \end{equation*}
\end{ex}

\begin{ex}
  Let $\vect{u}_1,\ldots,\vect{u}_k$ be vectors in $\R^n$, and let
  $S=\sspan\set{\vect{u}_1,\ldots,\vect{u}_k}$. Show that $S$ is the
  {\em smallest} subspace of $\R^n$ that contains
  $\vect{u}_1,\ldots,\vect{u}_k$.  Specifically, this means you have
  to show: if $V$ is any other subspace of $\R^n$ such that
  $\vect{u}_1,\ldots,\vect{u}_k\in V$, then $S\subseteq V$.
\end{ex}

