\section*{Exercises}

\begin{enumialphparenastyle}

\begin{ex}
  Use elementary operations to solve the system of equations
  \begin{equation*}
    \begin{array}{c}
      3x+y=3\\
      x+2y=1.
    \end{array}                                                                
  \end{equation*}

  \begin{sol}
    $(x,y) = (1,0)$.
  \end{sol}
\end{ex}

\begin{ex}
  Use elementary operations to find the point $\tup{x,y}$
  that lies on both lines $x+3y=1$ and $4x-y=3$.
  \begin{sol}
    $\begin{array}{c}
       x+3y=1 \\
       4x-y=3
     \end{array}
    $, Solution is: $(x,y) = \tup{\frac{10}{13},\frac{1}{13}}$
  \end{sol}
\end{ex}
 
\begin{ex}
  Use elementary operations to determine whether the three lines
  $x+2y=1$, $2x-y=1$, and $4x+3y=3$ have a common point of intersection.
  If so, find the point, and if not, tell why they don't have such a
  common point of intersection.
  \begin{sol}
    $\begin{array}{c}
       x+2y=1 \\
       2x-y=1 \\
       4x+3y=3
     \end{array}
    $, Solution is: $(x,y) = \tup{\frac{3}{5},\frac{1}{5}}$
  \end{sol}
\end{ex}

\begin{ex}
  Do the three planes, $x+y-3z=2$, $2x+y+z=1$, and $3x+2y-2z=0$ have a
  common point of intersection? If so, find one and if not, tell why
  there is no such point.
  \begin{sol}
    No solution exists. $
    \begin{array}{c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c}
      x &+& y &-& 3z &=& 2 \\
      2x &+& y &+& z &=& 1 \\
      3x &+& 2y &-& 2z &=& 0
    \end{array}
    $, after elementary operations: 
    $\begin{array}{c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c}
      x & &   &+& 4z &=& 0 \\
        & & y &-& 7z &=& 0 \\
        & &   & & 0 &=& 1
     \end{array}$.
    Thus one of the equations says $0=1$ in an equivalent system of
    equations.
  \end{sol}
\end{ex}

\begin{ex}
  Solve the following system of equations by back substitution.
  \begin{equation*}
    \begin{array}{c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c}
      x &+& 3y &-& 2z &=& 5 \\
        & & y  &+& 3z &=& 4 \\
        & &    & & z  &=& 1.
    \end{array}
  \end{equation*}

  \begin{sol}
    $(x,y,z) = (1,2,1)$.
  \end{sol}
\end{ex}

\begin{ex}
  Write the following system of linear equations as an augmented
  matrix. Caution: you first have to simplify and rearrange the
  equations so that ``like'' variables are lined up in columns. Write
  the variables in the order $x$, $y$, $z$.
  \begin{equation*}
    \begin{array}{c}
      x - 3z + 2y = 5 \\
      6 - x = 4 + y - z\\
      2x + 3 = x + 3y.
    \end{array}
  \end{equation*}

  \begin{sol}
    $
    \begin{array}{c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c}
      x &+& 2y &-& 3z &=& 5\\
      -x &-& y &+& z &=& -2\\
      x &-& 3y && &=& -3.\\
    \end{array}
    $
    As an augmented matrix:
    $
    \begin{mymatrix}{rrr|r}
      1 & 2 & -1 & 5\\
      -1 & -1 & 1 & -2\\
      1 & -3 & 0 & -3
    \end{mymatrix}.
    $
  \end{sol}
\end{ex}


\begin{ex}
Four times the weight of Gaston is 150kg more than the weight of
Ichabod. Four times the weight of Ichabod is 660kg less than seventeen
times the weight of Gaston. Four times the weight of Gaston plus the weight
of Siegfried equals 290kg. Brunhilde would balance all three of the
others. Find the weights of the four people.
\begin{sol}
$\begin{array}{c}
4g-I=150 \\
4I-17g=-660 \\
4g+s=290 \\
g+I+s-b=0
\end{array}
$, Solution is : $\set{g=60,I=90,b=200,s=50} $
\end{sol}
\end{ex}

\end{enumialphparenastyle}
