\section*{Exercises}

\begin{ex}
  Consider the set of $n\times n$ symmetric matrices, i.e., matrices
  satisfying $A=A^{T}$. Show that this set of symmetric matrices is a
  vector space and a subspace of the vector space of
  $n\times n$-matrices.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Consider the set of all vectors in $\R^{2},(x,y) $
  such that $x+y\geq 0$. Let the vector space operations be the usual ones. Is
  this a vector space? Is it a subspace of $\R^{2}$?
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Consider the vectors in $\R^{2},(x,y) $ such that $xy=0$. Is this a subspace of $\R^{2}$? Is it a vector space? The
  addition and scalar multiplication are the usual operations.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $V$ denote those polynomials $ax^2+bx+c\in\Poly_{2}$ such that
  $a+b+c=0$. Is $V$ a subspace of $\Poly_{2}$?  Explain.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $M,N$ be subspaces of a vector space $V$ and consider $M+N$
  defined as the set of all $m+n$ where $m\in M$ and $n\in N$. Show that $M+N$
  is a subspace of $V$.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $M,N$ be subspaces of a vector space $V$. Then $M\cap N$ consists
  of all vectors which are in both $M$ and $N$. Show that $M\cap N$ is a
  subspace of $V$.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $M,N$ be subspaces of a vector space $\R^{2}$. Then $N\cup
  M$ consists of all vectors which are in either $M$ or $N$. Show that $N\cup
  M $ is not necessarily a subspace of $\R^{2}$ by giving an example
  where $N\cup M$ fails to be a subspace.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex} Let $M=\set{\vect{u}=(u_{1},u_{2},u_{3},u_{4}) \in
\R^{4}:\abs{u_{1}} \leq 4}$. Is $M$ a subspace of $\R^4$?
\begin{sol}
This is not a subspace. $\begin{mymatrix}{r}
1 \\
1 \\
1 \\
1
\end{mymatrix} $ is in
it, but $20\begin{mymatrix}{r}
1 \\
1 \\
1 \\
1
\end{mymatrix} $ is not.
\end{sol}
\end{ex}

\begin{ex} Let $M=\set{\vect{u}=(u_{1},u_{2},u_{3},u_{4}) \in
\R^{4}:\sin (u_{1}) =1}$. Is $M$ a subspace of $\R^4$?
\begin{sol}
This is not a subspace.
\end{sol}
\end{ex}

\begin{ex} Let $W$ be a subset of $M_{22}$ given by
\[
W = \set{A \mid A \in M_{22}, A^T = A }
\]
In words, $W$ is the set of all symmetric $2 \times 2$-matrices. Is $W$ a subspace of $M_{22}$?
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $W$ be a subset of $M_{22}$ given by
\[
W = \set{\begin{mymatrix}{rr}
a  & b \\
c & d
\end{mymatrix} \mid a,b,c,d \in \R, a + b = c + d }
\]
Is $W$ a subspace of $M_{22}$?
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $W$ be a subset of $P_3$ given by
\[
W = \set{
ax^3 + bx^2 + cx + d \mid a,b,c,d \in \R, d = 0 }
\]
Is $W$ a subspace of $P_3$?
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $W$ be a subset of $P_3$ given by
\[
W = \set{
p(x) = ax^3 + bx^2 + cx + d \mid a,b,c,d \in \R, p(2) = 1 }
\]
Is $W$ a subspace of $P_3$?
%\begin{sol}
%\end{sol}
\end{ex}

