\section*{Exercises}

\begin{ex}
  Consider the set of symmetric $n\times n$-matrices%
  \index{matrix!symmetric}%
  \index{symmetric matrix}, i.e., matrices satisfying $A=A^{T}$. Show
  that this set of symmetric matrices is a subspace of $\Mat_{n,n}$,
  the vector space of $n\times n$-matrices.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Consider the set of all vectors
  $\begin{mysmallmatrix}{c}x\\y\end{mysmallmatrix}\in\R^2$ such that
  $x+y\geq 0$. Is this a subspace of $\R^{2}$?
  \begin{sol}
    No. It is not closed under scalar multiplication.
  \end{sol}
\end{ex}

\begin{ex}
  Consider the set of all vectors
  $\begin{mysmallmatrix}{c}x\\y\end{mysmallmatrix}\in\R^2$ such that $xy=0$. Is
  this a subspace of $\R^{2}$?
  \begin{sol}
    No. It is not closed under addition.
  \end{sol}
\end{ex}

\begin{ex}
  Let $V$ be the set of those polynomials $ax^2+bx+c\in\Poly_{2}$ such
  that $a+b+c=0$. Is $V$ a subspace of $\Poly_{2}$? Explain.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $U,W$ be subspaces of a vector space $V$ and consider $U+W$
  defined as the set of all vectors that can be written of the form
  $\vect{u}+\vect{w}$, where $\vect{u}\in U$ and $\vect{w}\in W$. Show
  that $U+W$ is a subspace of $V$.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $U,W$ be subspaces of a vector space $V$. Then $U\cap W$ consists
  of all vectors which are in both $U$ and $W$. Show that $U\cap W$ is a
  subspace of $V$.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $U,W$ be subspaces of a vector space $V$. Then $U\cup W$
  consists of all vectors which are in either $U$ or $W$. Show that
  $U\cup W$ is not necessarily a subspace of $V$ by giving an example
  where $U\cup W$ fails to be a subspace.
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $U=\set{\left.\mat{x,y,z}^T \in \R^{3}~\right\vert~\abs{x} \leq 4}$. Is $U$ a
  subspace of $\R^3$?
  \begin{sol}
    This is not a subspace.
    $\begin{mysmallmatrix}{r} 1 \\ 1 \\ 1 \end{mysmallmatrix}$ is in it, but
    $5\begin{mysmallmatrix}{r} 1 \\ 1 \\ 1 \end{mysmallmatrix} $ is not.
  \end{sol}
\end{ex}

\begin{ex}
  Let $W$ be the subset of\/ $\Mat_{2,2}$ given by
  \begin{equation*}
    W = \set{\left.\begin{mysmallmatrix}{rr}
        a  & b \\
        c & d
      \end{mysmallmatrix}~\right\vert~ a,b,c,d \in K,~ a + b = c + d }
  \end{equation*}
  Is $W$ a subspace of $\Mat_{2,2}$?
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $U=\set{
    A\in\Mat_{2,2}
    ~\left\vert~
      A\begin{mysmallmatrix}{rr} 1 & 0 \\ 1 & -1 \end{mysmallmatrix}
      = \begin{mysmallmatrix}{rr} 1 & 1 \\ 0 & -1 \end{mysmallmatrix} A
    \right.}$.
  Show that $U$ is a subspace of $\Mat_{2,2}$.
\end{ex}

\begin{ex}
  Let $W$ be the subset of\/ $\Poly_3$ given by
  \begin{equation*}
    W = \set{ ax^3 + bx^2 + cx + d \mid a,b,c,d \in K, d = 0 }
  \end{equation*}
  Is $W$ a subspace of $\Poly_3$?
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $W$ be the subset of\/ $\Poly_3$ given by
  \begin{equation*}
    W = \set{ p(x) \mid p(2) = 1 }
  \end{equation*}
  Is $W$ a subspace of $\Poly_3$?
  % \begin{sol}
  % \end{sol}
\end{ex}

\begin{ex}
  Let $W$ be the subset of $\Seq_{\R}$ consisting of all sequences
  that are alternating, i.e., where $a_i\geq 0$ for even $i$ and
  $a_i\leq 0$ for odd $i$, or vice versa. Is $W$ a subspace of
  $\Seq_{\R}$?
\end{ex}

\begin{ex}
  Let $W$ be the subset of $\Seq_{K}$ consisting of all sequences that
  satisfy the recurrence relation
  $a_{n+3} = a_{n+2} + 2a_{n+1} - a_{n}$. Is $W$ a subspace of
  $\Seq_{K}$?
\end{ex}

\begin{ex}\label{ex:periodic}
  A sequence $(a_i)_{i\in\N}$ is called \textbf{periodic}%
  \index{periodic sequence}%
  \index{sequence!periodic} if there exists some $k>0$ such that for
  all $i$, $a_{i+k}=a_i$. The number $k$ is called a \textbf{period}
  of the sequence. For example, the following is a periodic sequence
  with period $3$:
  \begin{equation*}
    1,5,-7,1,5,-7,1,5,-7,1,5,-7,1,\ldots
  \end{equation*}
  \begin{enumerate}
  \item Show that the set of all cyclic sequences of a fixed period
    $k$ forms a subspace of $\Seq_K$.
  \item More difficult: show that the set of all cyclic sequences of
    all periods forms a subspace of $\Seq_K$.
  \end{enumerate}
\end{ex}

\begin{ex}
  A function $f:\R\to\R$ is called \textbf{symmetric}%
  \index{function!symmetric}%
  \index{symmetric function} if $f(x)=f(-x)$ for all $x\in\R$.  Let
  $W$ be the subset of $\Func_{\R,\R}$ consisting of all symmetric
  functions. Is $W$ a subspace of $\Func_{\R,\R}$?
\end{ex}

\begin{ex}
  A function $f:\R\to\R$ is said to \textbf{vanish at infinity}%
  \index{function!vanishing at infinity}%
  \index{vanish at infinity} if $\lim_{x\to\infty}f(x)=0$ and
  $\lim_{x\to-\infty}f(x)=0$. Let $W$ be the subset of $\Func_{\R,\R}$
  consisting of all functions that vanish at infinity. Prove that $W$
  is a subspace of $\Func_{\R,\R}$.
\end{ex}

\begin{ex}
  Show that the sets $S=\set{x+2,(x+2)^2}$ and
  $T=\set{x^2-4, x^2+x-2}$ span the same subspace of $\Poly_2$.
\end{ex}

\begin{ex}
  Show that the sets $S=\set{
    \begin{mysmallmatrix}{rr} 1 & 0 \\ 0 & 1 \end{mysmallmatrix},~
    \begin{mysmallmatrix}{rr} 0 & 1 \\ 1 & 0 \end{mysmallmatrix}
  }$ and $T=\set{
    \begin{mysmallmatrix}{rr} 1 & 1 \\ 1 & -1 \end{mysmallmatrix},~
    \begin{mysmallmatrix}{rr} 1 & 1 \\ -1 & 1 \end{mysmallmatrix}
  }$
  span the same subspace of $\Mat_{2,2}$.
\end{ex}
