\Opensolutionfile{solutions}[ex]
\section*{Exercises}

\begin{enumialphparenastyle}

\begin{ex} Suppose you have $\R^{2}$ and the $+$ operation is as
follows:\ 
\begin{equation*}
\tup{a,b} +\tup{c,d} =\tup{a+d,b+c} .
\end{equation*}
Scalar multiplication is defined in the usual way. Is this a vector space?
Explain why or why not. 
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Suppose you have $\R^{2}$ and the $+$ operation is defined as
follows. 
\begin{equation*}
\tup{a,b} +\tup{c,d} =\tup{0,b+d}
\end{equation*}
Scalar multiplication is defined in the usual way. Is this a vector space?
Explain why or why not. 
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Suppose you have $\R^{2}$ and scalar multiplication is defined
as $c\tup{a,b} =\tup{a,cb} $ while vector addition is
defined as usual. Is this a vector space? Explain why or why not.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Suppose you have $\R^{2}$ and the $+$ operation is defined as
follows. 
\begin{equation*}
\tup{a,b} +\tup{c,d} =\tup{a-c,b-d}
\end{equation*}
Scalar multiplication is same as usual. Is this a vector space? Explain why
or why not. 
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} \label{functions}Consider all the functions defined on a non empty set
which have values in $\R$. Is this a vector space? Explain.
The operations are defined as follows. Here $f,g$ signify functions and $a$
is a scalar. 
\begin{eqnarray*}
\tup{f+g} \tup{x} &=&f\tup{x} +g\tup{x} \\
\tup{af} \tup{x} &=&a\tup{f\tup{x} }
\end{eqnarray*}
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} Denote by $\R^{\N}$ the set of real valued sequences.
For $\vect{a}\equiv \set{a_{n}} _{n=1}^{\infty },\vect{b}\equiv
\set{b_{n}} _{n=1}^{\infty }$ two of these, define their sum to be
given by 
\begin{equation*}
\vect{a}+\vect{b} =  \set{a_{n}+b_{n}} _{n=1}^{\infty }
\end{equation*}
and define scalar multiplication by 
\begin{equation*}
c\vect{a}=\set{ca_{n}} _{n=1}^{\infty }\text{ where }\vect{a}
=\set{a_{n}} _{n=1}^{\infty }
\end{equation*}
Is this a special case of Problem~\ref{functions}? Is this a vector space?
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $\C^{2}$ be the set of ordered pairs of complex numbers.
Define addition and scalar multiplication in the usual way.
\begin{equation*}
\tup{z,w} +\tup{\hat{z},\hat{w}} = \tup{z+\hat{z},w+
\hat{w}} ,\ u\tup{z,w} \equiv \tup{uz,uw}
\end{equation*}
Here the scalars are from $\C$. Show this is a vector space.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $V$ be the set of functions defined on a nonempty set which have
values in a vector space $W$. Is this a vector space? Explain.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the space of $m\times n$-matrices with operation of addition
and scalar multiplication defined the usual way. That is, if $A,B$ are two $
m\times n$-matrices and $c$ a scalar, 
\begin{equation*}
\tup{A+B} _{ij}=A_{ij}+B_{ij},\ \tup{cA} _{ij}\equiv c\tup{
A_{ij}}
\end{equation*}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the set of $n\times n$ symmetric matrices. That is, $A=A^{T}$.
In other words, $A_{ij}=A_{ji}$. Show that this set of symmetric matrices is
a vector space and a subspace of the vector space of $n\times n$-matrices.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the set of all vectors in $\R^{2},\tup{x,y} $
such that $x+y\geq 0$. Let the vector space operations be the usual ones. Is
this a vector space? Is it a subspace of $\R^{2}$?
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the vectors in $\R^{2},\tup{x,y} $ such that $xy=0$. Is this a subspace of $\R^{2}$? Is it a vector space? The
addition and scalar multiplication are the usual operations.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Define the operation of vector addition on $\R^{2}$ by $\tup{
x,y} +\tup{u,v} =\tup{x+u,y+v+1}$. Let scalar
multiplication be the usual operation. Is this a vector space with these
operations? Explain.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let the vectors be real numbers. Define vector space operations in the
usual way. That is $x+y$ means to add the two numbers and $xy$ means to
multiply them. Is $\R$ with these operations a vector space? Explain.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let the scalars be the rational numbers and let the vectors be
real numbers which are the form $a+b\sqrt{2}$ for $a,b$ rational numbers.
Show that with the usual operations, this is a vector space.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $\Poly_{2}$ be the set of all polynomials of degree 2 or
less. That is, these are of the form $a+bx+cx^{2}$. Addition is defined as 
\begin{equation*}
\tup{a+bx+cx^{2}} +\tup{\hat{a}+\hat{b}x+\hat{c}x^{2}}
=\tup{a+\hat{a}} +\tup{b+\hat{b}} x+\tup{c+\hat{c}}
x^{2}
\end{equation*}
and scalar multiplication is defined as 
\begin{equation*}
d\tup{a+bx+cx^{2}} =da+dbx+cdx^{2}
\end{equation*}
Show that, with this definition of the vector space operations that $\Poly_{2}$ is a vector space. Now let $V$ denote those polynomials $a+bx+cx^{2}$
such that $a+b+c=0$. Is $V$ a subspace of $\Poly_{2}$? Explain.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $M,N$ be subspaces of a vector space $V$ and consider $M+N$
defined as the set of all $m+n$ where $m\in M$ and $n\in N$. Show that $M+N$
is a subspace of $V$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $M,N$ be subspaces of a vector space $V$. Then $M\cap N$ consists
of all vectors which are in both $M$ and $N$. Show that $M\cap N$ is a
subspace of $V$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $M,N$ be subspaces of a vector space $\R^{2}$. Then $N\cup
M$ consists of all vectors which are in either $M$ or $N$. Show that $N\cup
M $ is not necessarily a subspace of $\R^{2}$ by giving an example
where $N\cup M$ fails to be a subspace.
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} \label{4-july-prob1}Let $X$ consist of the real valued functions which
are defined on an interval $\mat{a,b}$. For $f,g\in X,\;f+g$ is the
name of the function which satisfies $\tup{f+g} \tup{x}
=f\tup{x} +g\tup{x}$. For $s$ a real number, $
\tup{s f} \tup{x} = s \tup{f\tup{x}
}$. Show this is a vector space. 
\begin{sol}
The axioms of a vector space all hold because they
hold for a vector space. The only thing left to verify is the
assertions about the things which are supposed to exist. $0$ would
be the zero function which sends everything to $0$. This is an additive
identity. Now if $f$ is a function, $-f\tup{x} \equiv \tup{
-f\tup{x} }$. Then
\[
\tup{f+\tup{-f} } \tup{x} \equiv f\tup{x}
+\tup{-f} \tup{x} \equiv f\tup{x} +\tup{-f\tup{
x} } =0
\]
Hence $f+-f=\mathbf{0}$. For each $x\in \mat{a,b}$, let $%
f_{x}\tup{x} =1$ and $f_{x}\tup{y} =0$ if $y\neq x$. Then
these vectors are obviously linearly independent.
\end{sol}
\end{ex}

\begin{ex} Consider functions defined on $\set{1,2,\cdots ,n} $ having
values in $\R$. Explain how, if $V$ is the set of all such
functions, $V$ can be considered as $\R^{n}$.
\begin{sol}
Let $f\tup{i} $ be the $i^{th}$ component of a vector $
\vect{x}\in \R^{n}$. Thus a typical element in $\R^{n}$ is $
\tup{f\tup{1} ,\cdots ,f\tup{n} }$.
\end{sol}
\end{ex}

\begin{ex} Let the vectors be polynomials of degree no more than 3. Show that
with the usual definitions of scalar multiplication and addition wherein,
for $p\tup{x} $ a polynomial, $\tup{a p} \tup{
x} = a p\tup{x} $ and for $p,q$ polynomials $\tup{
p+q} \tup{x} =  p\tup{x} +q\tup{x}$, this
is a vector space.
\begin{sol}
This is just a subspace of the vector space of functions
because it is closed with respect to vector addition and scalar
multiplication. Hence this is a vector space.
\end{sol}
\end{ex}

\end{enumialphparenastyle}
