\thispagestyle{plain}


\setlength{\parskip}{0pt}

\begin{center}
  \fontsize{14pt}{16pt}\selectfont\textcolor{titletextcolour}{\textbf{Revision history \\ \smallskip  Current revision: Dal 2018 A}}
  \\\bigskip\fontsize{10pt}{12pt}\selectfont\textcolor{titletextcolour}{This
    printing: version {\revid} of {\revdate}}
\end{center}

{\footnotesize

  \begin{center}
    Extensive edits, additions, and revisions have been completed by
    Peter Selinger and other contributors. \\
    Prior extensive edits, additions, and revisions were made by the
    editorial staff at Lyryx Learning. \\
    All new content (text and images) is released under the same
    license as noted above.
  \end{center}

  %% with package enuitem can use [leftmargin=*] after \begin{itemize}
  \begin{tabu}{p{5em} X[j,m]} %\hline
    % Revision & \centering{Author: Significant Changes} \\
    \hline
    Dal 2018 A &
    \begin{itemize}
    \item P. Selinger: updated title and front matter.
    \item P. Selinger: Extensive revisions and additions. Rewrote most
      of Chapters 9--11 from scratch. Added application on perspective
      rendering, error correcting codes, least squares approximations
      and curve fitting, an audio demo for Fourier series,
      sketching of quadratic forms, and principal component analysis.
    \end{itemize}
    \\ \hline
    Dal 2017 B &
    \begin{itemize}
    \item P. Selinger, M.B. Langlois: Re-ordered chapters. Extensive
      revisions to Chapters 1--8. Added new sections on fields,
      cryptography, geometric interpretation of linear
      transformations, recurrences, systems of linear differential
      equations.
    \end{itemize}
    \\ \hline
    2017 A &
    \begin{itemize}
    \item Lyryx: Front matter has been updated including cover,
      copyright, and revision pages.
    \item I. Farah: contributed edits and revisions, particularly the
      proofs in the Properties of
      Determinants II: Some Important Proofs section.
    \end{itemize}
    \\ \hline
    2016 B &
    \begin{itemize}
    \item Lyryx: The text has been updated with the addition of
      subsections on Resistor Networks and the Matrix Exponential
      based on original material by K. Kuttler.
    \item Lyryx: New example on Random Walks developed.
    \end{itemize}
    \\ \hline
    2016 A &
    \begin{itemize}
    \item Lyryx: The layout and appearance of the text has been
      updated, including the title page and newly designed back
      cover.
    \end{itemize}
    \\ \hline
    2015 A &
    \begin{itemize}
    \item Lyryx: The content was modified and adapted with the
      addition of new material and several images throughout.
    \item Lyryx: Additional examples and proofs were added to existing
      material throughout.
    \end{itemize}
    \\ \hline
    2012 A &
    \begin{itemize}
    \item \textbf{Original text} by K. Kuttler of Brigham Young
      University. That version is used under Creative Commons license
      CC BY (\url{https://creativecommons.org/licenses/by/3.0/}) made
      possible by funding from The Saylor Foundation's Open Textbook
      Challenge. See
      \href{https://www.saylor.org/site/wp-content/uploads/2012/02/Elementary-Linear-Algebra-1-30-11-Kuttler-OTC.pdf}{Elementary Linear Algebra}
      for more information and the original version.
    \end{itemize}
    \\ \hline
  \end{tabu}
  \medskip
}

\setlength{\parskip}{\baselineskip}


