\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface} %Adds preface to the table of contents

\textit{\bookfulltitle} presents an introduction to the fascinating
subject of linear algebra. It is intended for students in the first or
second year of university, and contains enough material for a
2-semester course. Major topics of linear algebra are presented in
detail, and many applications are given. Although it is not a
proof-oriented book, proofs of most important theorems are
provided. 

Each section begins with a list of desired outcomes which a student
should be able to achieve upon completing the chapter.  Throughout the
text, examples and diagrams are given to reinforce ideas and provide
guidance on how to approach various problems. Students are encouraged
to work through the suggested exercises provided at the end of each
section. Selected solutions to these exercises are given at the end of
the text.

\subsection*{Open source}

This is an open-source textbook, licensed under the Creative Commons
``CC~BY~4.0'' license. This means, among other things, that you are
permitted to copy and redistribute this textbook in any medium or
format. For example, you can download this textbook for free, print
copies for yourself or others, or share it on the internet.

You are also permitted to make changes to this textbook and distribute
the changed version, provided that you follow the terms of the license
(for example, you must give appropriate credit to previous authors,
and you must distribute your changes under the same open-source
license). If you are an instructor who wants to use this textbook in a
course, you can add new sections, change the notation, or add more
examples and exercises. If you make revisions, please send them to me
so that I can consider incorporating them in future versions of this
book. Please see \url{https://creativecommons.org/licenses/by/4.0/}
for details of the licensing terms.

This textbook has a website at
\url{https://www.mathstat.dal.ca/~selinger/linear-algebra/}.  There,
you can find the most up-to-date version. The website also contains
some supplementary material, a link to the source code and license,
options for purchasing a printed version of this book, and more.

\subsection*{Reporting typos}

Like all books, this book likely contains some typos and other
errors. However, since it is an open-source book, typos can easily be
fixed and an updated version posted online. It is my intention to fix
all typos. If you find a typo (no matter how small), please report it
to me at
\href{mailto:selinger@mathstat.dal.ca}{selinger@mathstat.dal.ca}. Thanks
to the following people who have already reported typos: Yaser
Alkayale, Robert Earle, Melissa Huggan, Marie-Andr\'ee Langlois,
Brenda Le, Sarah Li, Arman Kerimbek, Li Wei Men, and Dallas Sawtell.
