\section{Parametric lines}

\begin{outcome}

\begin{enumerate}

\item[A.] Find the vector and parametric equations of a line.

\end{enumerate}
\end{outcome}

We can use the concept of vectors and points to find equations for arbitrary lines in $\R^n$, although in this section the focus will be on lines in $\R^3$. 

To begin, consider the case $n=1$ so we have $\R^{1}=\R$. There is only one line here which is the familiar number line, that is $\R$ itself. Therefore it is not necessary to explore the case of $n=1$ further. 

Now consider the case where $n=2$, in other words $\R^2$.  Let $P $ and $P_0$ be
two different points in $\R^{2}$ which are contained in a line
$L$. Let $\vect{p}$ and $\vect{p}_0$ be the position vectors for the points $P$ and $P_0$
respectively. Suppose that $Q$ is an arbitrary point on $L$. Consider the following
diagram.

\begin{center}
\begin{tikzpicture}
\draw[<->,thick,blue](-0.5,-0.5)--(2.5,2.5);
\draw[help lines](0,0)--(1,0)--(1,1);
\draw[help lines](0,0)--(2,0)--(2,2);
\node[above left] at (0,0){$P_0$};
\node[above left] at (1,1){$P$};
\node[above left] at (2,2){$Q$};
\end{tikzpicture}
\end{center}

Our goal is to be able to define $Q$ in terms of $P$  and $P_0$. Consider the vector $\longvect{P_0P} = \vect{p} - \vect{p}_0$ which
has its tail at $P_0$ and point at $P$. If we add $\vect{p} - \vect{p}_0$ to the position vector $\vect{p}_0$ for $P_0$,  
the sum would be a vector with its point at $P$. 
In other words,
\begin{equation*}
\vect{p} = \vect{p}_0 + (\vect{p} - \vect{p}_0)
\end{equation*}

Now suppose we were to add $t(\vect{p} - \vect{p}_0)$ to $\vect{p}$ where $t$ is some scalar.
You can see that by doing so, we could find a vector with its point at $Q$. In other words, we can find $t$ such that 
\begin{equation*}
\vect{q} = \vect{p}_0 + t \tup{\vect{p}- \vect{p}_0}
\end{equation*}

This equation determines the line $L$ in $\R^2$. In fact, it determines a line $L$ in $\R^n$. Consider the following definition.

\begin{definition}{Vector equation of a line}{vectorequationofline}
Suppose a line $L$ in $\R^{n}$ contains the two\index{lines!vector equation} different points $P$ and 
$P_0$. Let $\vect{p}$ and $\vect{p}_0$ be the position vectors of these two points, respectively.
Then, $L$ is the collection of points $Q$ which have the position vector $\vect{q}$ given by
\begin{equation*}
\vect{q}=\vect{p}_0+t\tup{\vect{p}-\vect{p}_0}
\end{equation*}
where $t\in \R$. 

Let $\vect{d} = \vect{p} - \vect{p}_0$. Then $\vect{d}$ is the \textbf{direction vector for $L$}\index{direction vector} and the \textbf{vector equation} for $L$ is given by 
\begin{equation*}
\vect{p}=\vect{p}_0+t\vect{d}, t\in\R
\end{equation*}
\end{definition}

Note that this definition agrees with the usual notion of a
line in two dimensions and so this is consistent with earlier concepts. Consider now points in $\R^3$. If a point $P \in \R^3$ is given by $P = \tup{x,y,z }$, $P_0 \in \R^3$ by $P_0 = \tup{x_0, y_0, z_0 }$, then we can write
\begin{equation*}
\begin{mymatrix}{c}
x \\
y \\
z 
\end{mymatrix} = 
\begin{mymatrix}{c}
x_0 \\
y_0 \\
z_0 
\end{mymatrix}
+
t
\begin{mymatrix}{c}
a \\
b \\
c 
\end{mymatrix}
\end{equation*}
where $\vect{d} = \begin{mymatrix}{c}
a \\
b \\
c 
\end{mymatrix}$. This is the vector equation of $L$ written in \textbf{component form}\index{component form}.

The following theorem claims that such an equation is in fact a line. 

\begin{proposition}{Algebraic description of a straight line}{algebraicstraightline}
Let $\vect{a},\vect{b}\in \R^{n}$ with $\vect{b}\neq \vect{0}$. 
Then $\vect{x}=\vect{a}+t\vect{b},\; t\in \R$, is a
line.
\end{proposition}

\begin{proof}
Let $\vect{x}_1, \vect{x}_2 \in \R^n$. 
Define $\vect{x}_1=\vect{a}$ and let $\vect{x}_2-\vect{x}_1=\vect{b}$.
Since $\vect{b} \neq \vect{0}$, it follows that $\vect{x}_2\neq \vect{x}_1.$ Then
$\vect{a}+t\vect{b}=\vect{x}_1 + t\tup{\vect{x}_2-\vect{x}_1} $. It follows that  
$\vect{x}=\vect{a}+t\vect{b}$ is a line containing the two different points 
$X_1$ and $X_2$ whose position vectors are given by $\vect{x}_1$ and $\vect{x}_2$ respectively. 
\end{proof}

We can use the above discussion to find the equation of a line when given two distinct points.
Consider the following example. 

\begin{example}{A line from two points}{linefromtwopoints}
Find a vector equation for the line through the points $P_0 = \tup{
1,2,0} $ and $P = \tup{2,-4,6}.$
\end{example}

\begin{solution}
We will use the definition of a line given above in Definition \ref{def:vectorequationofline} to write
this line in the form 
\begin{equation*}
\vect{q}=\vect{p}_0+t\tup{\vect{p}-\vect{p}_0}
\end{equation*}
Let $\vect{q} = 
\begin{mymatrix}{c}
x \\
y \\
z
\end{mymatrix}
$. Then, we can find $\vect{p}$ and $\vect{p}_0$ by taking the position vectors of points $P$ and $P_0$ 
respectively. 
Then, 
\begin{equation*}
\vect{q}=\vect{p}_0+t\tup{\vect{p}-\vect{p}_0}
\end{equation*}
can be written as
\begin{equation*}
\begin{mymatrix}{c}
x \\
y \\
z \\
\end{mymatrix}
 =
\begin{mymatrix}{c}
1 \\
2 \\
0
\end{mymatrix}
+
t
\begin{mymatrix}{r}
 1 \\
-6 \\
6  
\end{mymatrix},
\;t\in
\R
\end{equation*}
Here, the direction vector $
\begin{mymatrix}{r}
1 \\
-6 \\
6 
\end{mymatrix}
$
is obtained by $
\vect{p} - \vect{p}_0 = 
\begin{mymatrix}{r}
2 \\
-4 \\
6
\end{mymatrix}
-
\begin{mymatrix}{r}
1 \\
2 \\
0
\end{mymatrix}
$
as indicated above in Definition \ref{def:vectorequationofline}.
\end{solution} 

Notice that in the above example we said that we found ``a'' vector
equation for the line, not ``the'' equation.  The reason for this
terminology is that there are infinitely many different vector 
equations for the same line. To see this, replace $t$ with another
parameter, say $3s.$ Then you obtain a different vector equation for the same
line because the same set of points is obtained.

In Example \ref{exa:linefromtwopoints}, the vector given by 
$\begin{mymatrix}{r}
 1 \\
-6 \\
6 
\end{mymatrix}$
is the direction vector defined in Definition \ref{def:vectorequationofline}.
If we know the direction vector of a line, as well as a point on the line,
we can find the vector equation. 

Consider the following example. 

\begin{example}{A line from a point and a direction vector}{linepointanddirectionvector}
Find a vector equation for the line which contains the point $P_0 = \tup{
1,2,0} $ and has direction vector $\vect{d} = 
\begin{mymatrix}{c}
1 \\
2 \\
1
\end{mymatrix}
$
\end{example}

\begin{solution}
We will use Definition \ref{def:vectorequationofline} to write this line in the form 
$\vect{p}=\vect{p}_0+t\vect{d},\; t\in \R$. We are given the direction vector $\vect{d}$. 
In order to find $\vect{p}_0$, we can use the position vector of the point $P_0$. 
This is given by 
$\begin{mymatrix}{c}
1 \\
2 \\
0
\end{mymatrix}.
$
Letting $\vect{p}
=
\begin{mymatrix}{c}
 x \\
y \\
z
\end{mymatrix}
$,
the equation for the line is given by 
\begin{equation}
\begin{mymatrix}{c}
x \\
y \\
z
\end{mymatrix}
=
\begin{mymatrix}{c}
1 \\
2 \\
0
\end{mymatrix}
+ 
t
\begin{mymatrix}{c}
1 \\
2 \\
1
\end{mymatrix},
\;t\in
\R \label{vectoreqn}
\end{equation}
\end{solution}

We sometimes elect to write a line such as the one given in \ref{vectoreqn} in the form
\begin{equation}
\begin{array}{ll}
\left. \begin{array}{l}
x=1+t \\
y=2+2t \\
z=t 
\end{array} \right\} & 
\mbox{where} \; t\in \R 
\end{array}
\label{parameqn}
\end{equation}
This set of equations give the same information as \ref{vectoreqn}, and is called the \textbf{parametric equation of the line}. 

Consider the following definition. 

\begin{definition}{Parametric equation of a line}{parametricequation}
Let $L$ be a line in $\R^3$ which has direction vector $\vect{d} = \begin{mymatrix}{c}
a \\
b \\
c
\end{mymatrix}$
and goes through the point $P_0 = \tup{x_0, y_0, z_0 }$.
Then, letting $t$ be a parameter, we can write $L$ as 
\begin{equation*}
\begin{array}{ll}
\left.
\begin{array}{c}
x = x_0 + ta \\
y = y_0 + tb \\
z = z_0 + tc
\end{array}
\right\} & 
\mbox{where} \; t\in \R 
\end{array}
\end{equation*}
This is called a \textbf{parametric equation}\index{lines!parametric equation} of the line $L$.
\end{definition}

You can verify that the form discussed following Example \ref{exa:linepointanddirectionvector} in equation  \ref{parameqn} is of the form
given in Definition \ref{def:parametricequation}.

There is one other form for a line which is useful, which is the \textbf{symmetric form}. 
Consider the line given by  \ref{parameqn}. You can
solve for the parameter $t$ to write
\begin{equation*}
\begin{array}{l}
t=x-1 \\
t=\frac{y-2}{2} \\
t=z
\end{array}
\end{equation*}
Therefore, 
\begin{equation*}
x-1=\frac{y-2}{2}=z
\end{equation*}
This is the \textbf{symmetric form} of the line.\index{lines!symmetric form}

In the following example, we look at how to take the equation of a line from symmetric form to 
parametric form.

\begin{example}{Change symmetric form to parametric form}{symmetrictoparametric}
Suppose the
\textbf{symmetric form of a line} is
\begin{equation*}
\frac{x-2}{3}=\frac{y-1}{2}=z+3
\end{equation*}
Write the line in parametric form as well as vector form.
\end{example}

\begin{solution}
We want to write this line in the form given by Definition \ref{def:parametricequation}. This is of the form 
\begin{equation*}
\begin{array}{ll}
\left.
\begin{array}{c}
x = x_0 + ta \\
y = y_0 + tb \\
z = z_0 + tc
\end{array}
\right\} & 
\mbox{where} \; t\in \R 
\end{array}
\end{equation*}

Let $t=\frac{x-2}{3},t=\frac{y-1}{2}$ and $t=z+3$, as given in the symmetric form of the line.
 Then solving for $x,y,z,$
yields
\begin{equation*}
\begin{array}{ll}
\left.
\begin{array}{c}
x=2 + 3t \\
y=1 + 2t \\
z=-3 + t 
\end{array}
\right\} & 
\mbox{with} \;t\in \R
\end{array}
\end{equation*}

This is the parametric equation for this line. 

Now, we want to write this line in the form given by Definition \ref{def:vectorequationofline}.
This is the form 
\begin{equation*}
\vect{p}=\vect{p}_0+t\vect{d}
\end{equation*}
where $t\in \R$.
This equation becomes
\begin{equation*}
\begin{mymatrix}{c}
x \\
y \\
z
\end{mymatrix} =
\begin{mymatrix}{r}
2 \\
1 \\
-3 
\end{mymatrix}
+
t
\begin{mymatrix}{r}
3 \\
2 \\
1 
\end{mymatrix},
\;t\in
\R
\end{equation*}
\end{solution}

% ----------------------------------------------------------------------
Another application of the geometric description of the dot product is
in finding the angle between two lines. Typically one would assume that the
lines intersect. In some situations, however, it may make sense to ask this
question when the lines do not intersect, such as the angle between
two object trajectories. In any case we understand it to mean the
smallest angle between (any of) their direction vectors. The only
subtlety here is that if $\vect{u}$ is a direction vector for a line,
then so is any multiple $k\vect{u}$, and thus we will find complementary angles
among all angles between direction vectors for two lines, and we
simply take the smaller of the two.

\begin{example}{Find the angle between two lines}{anglebetweentwolines}
Find the angle between the two lines
\begin{equation*}
L_1:  \;
\begin{mymatrix}{r}
x \\
y \\
z 
\end{mymatrix}
 = 
\begin{mymatrix}{r}
1 \\
2 \\
0
\end{mymatrix} +t\begin{mymatrix}{r}
-1 \\
1 \\
2
\end{mymatrix} 
\end{equation*}
 and
\begin{equation*}
L_2: \;
\begin{mymatrix}{r}
x \\
y \\
z
\end{mymatrix}
 = 
\begin{mymatrix}{r}
0 \\
4 \\
-3
\end{mymatrix}
 +s\begin{mymatrix}{r}
2 \\
1 \\
-1
\end{mymatrix}
\end{equation*}
\end{example}

\begin{solution}
You can verify that these lines do not intersect, but as discussed
above this does not matter and we simply find the smallest angle
between any directions vectors for these lines.

To do so  we first find the angle
between the direction vectors given above:
\begin{equation*}
\vect{u}=\begin{mymatrix}{r}
-1 \\
1 \\
2
\end{mymatrix},\;
\vect{v}=\begin{mymatrix}{r}
2 \\
1 \\
-1
\end{mymatrix}
\end{equation*}

In order to find the angle, we solve the following equation for $\theta$
\begin{equation*}
\vect{u}\dotprod \vect{v}=\norm{\vect{u}} \norm{\vect{v}
} \cos \theta
\end{equation*}
to obtain $\cos \theta = -\frac{1}{2}$ and since we choose included
angles between $0$ and $\pi$ we obtain $\theta = \frac{2 \pi}{3}$.


Now the angles between any two direction vectors for these lines will
either be $\frac{2 \pi}{3}$ or its complement $ \phi = \pi -  \frac{2 \pi}{3}
= \frac{\pi}{3}$. We choose the smaller angle, and therefore conclude that the angle between the two lines is $\frac{\pi}{3}$.
\end{solution}

% ----------------------------------------------------------------------

We will conclude this section with an important application of projections. Suppose a line $L$ and a point $P$ are given such that $P$ is not contained in $L$. Through the use of projections, we can determine the shortest distance from $P$ to $L$. 

\begin{example}{Shortest distance from a point to a line}{shortestpointline}
Let $P = (1,3,5)$ be a point in $\R^3$, and let $L$ be the line which goes through point $P_0 = (0,4,-2)$ with direction vector $\vect{d} = \begin{mymatrix}{r}
2 \\
1 \\
2
\end{mymatrix}
$.  Find the shortest distance from $P$ to the line $L$, and find the point $Q$ on $L$ that is closest to $P$. 
\end{example}

\begin{solution}
In order to determine the shortest distance from $P$ to $L$, we will first find the vector $\longvect{P_0P}$ and then find the projection of this vector onto $L$. 
The vector $\longvect{P_0P}$ is given by 
\[
\longvect{P_0P}=
\begin{mymatrix}{r}
1 \\
3 \\
5
\end{mymatrix}
-
\begin{mymatrix}{r}
0 \\
4 \\
-2
\end{mymatrix}
 = \begin{mymatrix}{r}
1 \\
-1 \\
7
\end{mymatrix}
\]

Then, if $Q$ is the point on $L$ closest to $P$, it follows that 
\begin{eqnarray*}
\longvect{P_0Q} &=& \func{proj}_{\vect{d}}\longvect{P_0P} \\
&=& \tup{\frac{ \longvect{P_0P}\bullet \vect{d}}{\norm{\vect{d}}^2}} \vect{d} \\
&=& 
\frac{15}{9} \begin{mymatrix}{r}
2 \\
1 \\
2
\end{mymatrix} \\
&=&
\frac{5}{3} \begin{mymatrix}{r}
2 \\
1 \\
2
\end{mymatrix}
\end{eqnarray*}

Now, the distance from $P$ to $L$ is given by 
\[
\norm{\longvect{QP}} = \norm{\longvect{P_0P} - \longvect{P_0Q}}
 = \sqrt{26} 
\]

The point $Q$ is found by adding the vector $\longvect{P_0Q}$ to the position vector $\longvect{0P_0}$ for $P_0$ as follows
\begin{eqnarray*}
\begin{mymatrix}{r}
0 \\
4 \\
-2
\end{mymatrix}
+
\frac{5}{3}
\begin{mymatrix}{r}
2 \\
1 \\
2
\end{mymatrix} 
&=& 
\begin{mymatrix}{r}
\vspace{0.05in}\frac{10}{3} \\
\vspace{0.05in}\frac{17}{3} \\
\vspace{0.05in}\frac{4}{3}
\end{mymatrix}
\end{eqnarray*}

Therefore, $Q = (\frac{10}{3}, \frac{17}{3}, \frac{4}{3})$. 

\end{solution}
