\section{Algebraic view of systems of equations}

\begin{outcome}
  \begin{enumerate}
  \item Recognize the difference between a linear equation and a
    non-linear equation.
  \item Determine whether a tuple of real numbers is a solution for a
    system of linear equations.
  \item Understand what it means for a system of linear equations to
    be consistent or inconsistent.
  \end{enumerate}
\end{outcome}

We have taken an in-depth look at graphical representations of systems of equations, as well as how to find possible
solutions graphically. Our attention now turns to working with systems algebraically.

\begin{definition}{Linear equation}{linear-equation}
A \textbf{linear equation}\index{linear equation} is an equation of
the form
\begin{equation*}
  a_1x_1 + a_2x_2 + \ldots + a_nx_n = b.
\end{equation*}
Here, $a_1,\ldots,a_n$ are real numbers called the
\textbf{coefficients}\index{coefficient} of the equation, $b$ is a
real number called the \textbf{constant term}\index{constant term}
of the equation, and $x_1,\ldots,x_n$ are
\textbf{variables}\index{variable}.
\end{definition}

Real numbers, such as the coefficients $a_1,\ldots,a_n$, the constant
term $b$, or the values of the variables $x_1,\ldots,x_n$, will also
be called \textbf{scalars}\index{scalar}. For now, the word
``scalar'' is just a synonym for ``real number''. Later, in
Section~\ref{sec:fields}, we will discover other kinds of scalars.

\begin{example}{Linear vs. non-linear equation}{linear-vs-non-linear}
  Which of the following equations are linear?
  \begin{equation*}
    \begin{array}{c}
      2x+3y=5 \\
      2x^2+3y=5 \\
      2\sqrt{x} + 3y = 5 \\
      (\sqrt{2}) x + 3y = 5^2
    \end{array}
  \end{equation*}
\end{example}

\begin{solution}
  The equation $2x+3y=5$ is linear. The equation $2x^2+3y=5$ is not
  linear, because it contains the square of a variable instead of a
  variable.  The equation $2\sqrt{x} + 3y = 5$ is also not linear,
  because the square root is applied to one of the variables. On the
  other hand, the equation $(\sqrt{2}) x + 3y = 5^2$ is linear,
  because $\sqrt{2}$ and $5^2$ are real numbers, and can
  therefore be used as coefficients and constant terms.
\end{solution}

We also permit minor notational variants of linear equations. The
equation $2x-3y=5$ is linear although
Definition~\ref{def:linear-equation} does not mention subtraction,
because it can be regarded as just another notation for
$2x+(-3)y = 5$. Similarly, the equation $2x=5+3y$ can be regarded as
linear, because it can be easily rewritten as $2x-3y=5$ by bringing
all the variables (and their coefficients) to the left-hand side.
When we need to emphasize that some linear equation is literally of
the form $a_1x_1 + a_2x_2 + \ldots + a_nx_n = b$, we say that the
equation is in \textbf{standard form}%
\index{linear equation!standard form}%
\index{standard form!of a linear equation}. Thus, the standard form of
the equation $2x=5+3y$ is $2x+(-3)y=5$.

A \textbf{solution}\index{linear equation!solution} to a linear
equation is an assignment of real numbers to the variables, making the
equation true. More precisely, if $r_1,\ldots,r_n$ are real numbers,
the assignment $x_1=r_1$, \ldots, $x_n=r_n$ is a solution to the
equation in Definition~\ref{def:linear-equation} if the real number
$a_1r_1 + a_2r_2 + \ldots + a_nr_n$ is equal to the real number
$b$. To save space, we often write solutions in \textbf{tuple
  notation}\index{tuple}\footnote{The terminology ``tuple'' arose as
  follows. A collection of two items is called a ``pair'', a
  collection of three items is called a ``triple'', followed by
  ``quadruple'', ``quintuple'', ``sextuple'', and so on. You have to
  know Latin to know what the next ones are called. To avoid these
  Latin terms, mathematicians started saying $4$-tuple, $5$-tuple,
  $6$-tuple and so on, and more generally, $n$-tuple for an ordered
  collection of $n$ items. When $n$ doesn't matter or is clear from
  the context, we often just say ``tuple''.}  as
$(x_1,\ldots,x_n) = (r_1,\ldots,r_n)$. When there is no doubt about
the order of the variables, we also often simply write the solution as
$(r_1,\ldots,r_n)$.

\begin{example}{Solutions of a linear equation}{linear-equation-solutions}
  Consider the linear equation $2x+3y-4z=5$. Which of the following
  are solutions? (a) $(x,y,z)=(1,1,0)$, (b) $(x,y,z)=(0,3,1)$, (c)
  $(x,y,z)=(1,1,1)$.
\end{example}

\begin{solution}
  The assignment $(x,y,z) = (1,1,0)$ is a solution because
  $2(1)+3(1)-4(0) = 5$. The assignment $(x,y,z) = (0,3,1)$ is also a
  solution, because $2(0) + 3(3) - 4(1) = 5$. On the other hand,
  $(x,y,z) = (1,1,1)$ is not a solution, because
  $2(1) + 3(1)-4(1) = 1 \neq 5$.
\end{solution}

A system of linear equations is just several linear equations taken together.

\begin{definition}{System of linear equations}{system-of-linear-equations}
A \textbf{system of linear equations}\index{system of linear equations} is a list of equations
\begin{equation*}
\begin{array}{c@{~}c@{~}c}
a_{11}x_1+a_{12}x_2+\ldots+a_{1n}x_n&=&b_1 \\
a_{21}x_1+a_{22}x_2+\ldots+a_{2n}x_n&=&b_2 \\
\vdots \\
a_{m1}x_1+a_{m2}x_2+\ldots+a_{mn}x_n&=&b_m,
\end{array}
\end{equation*}
where $a_{ij}$ and $b_j$ are scalars (i.e., real numbers). The above
is a system of $m$ equations in the $n$ variables,
$x_1,x_2\ldots,x_n$.  As before, the numbers $a_{ij}$ are
called the \textbf{coefficients}\index{coefficient} and the numbers
$b_j$ are called the \textbf{constant terms}\index{constant term} of
the system of equations.
\end{definition}

The relative size of $m$ and $n$ is not important here. We may have
more variables than equations, more equations than variables, or an
equal number of equations and variables.

A \textbf{solution}\index{system of linear equations!solution} to a
system of linear equations is an assignment of real numbers to the
variables that is a solution to {\em all} of the equations in the
system.

\begin{example}{Solutions of a system of linear equations}{linear-system-solutions}
  Consider the system of linear equations
  \begin{equation*}
    \begin{array}{c@{~}c@{~}c@{~}c@{~}c@{~}c@{~}c}
      2x&+&3y&-&4z &=& 5 \\
      -2x&+&y&+&2z &=& -1.
    \end{array}
  \end{equation*}
  Which of the following are solutions of the system? (a)
  $(x,y,z)=(1,1,0)$, (b) $(x,y,z)=(6,3,4)$, (c) $(x,y,z)=(0,3,1)$.
\end{example}

\begin{solution}
  The assignment $(x,y,z)=(1,1,0)$ is a solution of this system of
  equations, because it is a solution to the first equation and the
  second equation. Also, $(x,y,z)=(6,3,4)$ is another solution of this
  system of equations (check this!).  On the other hand,
  $(x,y,z)=(0,3,1)$ is not a solution of the system, because although
  it is a solution to the first equation, it is not a solution to the
  second equation.
\end{solution}

Recall from Section~\ref{sec:systems-geometric} that a system of
equations either has a unique solution, infinitely many solutions, or
no solution. It is very important to us whether a system of equations
has solutions or not. For this reason, we introduce the following terminology:

\begin{definition}{Consistent and inconsistent systems}{consistent-and-inconsistent}
A system of linear equations is called
\textbf{consistent}\index{consistent system}\index{system of linear equations!consistent}
if there exists at least one solution. It is called
\textbf{inconsistent}\index{inconsistent system}\index{system of linear equations!inconsistent}
if there is no solution.
\end{definition}

If we think of each equation as a condition that must be satisfied by
the variables, consistent means that there is some choice of values
for the variables which can satisfy \textbf{all} of the
conditions. Inconsistent means that there is no such choice of values
for the variables. In the following sections, you will learn a method
for determining whether a system of equations is consistent or not,
and in case it is consistent, to find all of its solutions.

